Reproducible Research Project 1
============================================

##Part 1: Loading and processing the data  
The data set contains 2 months (Oct-Nov 2012) of recorded steps each day, broken down into 5 minute intervals. This section completes the following:

- Loads the necessary R packages
- Reads in the data
- Formats the date column as class "date"
- Print out the first few lines of data  
  
####The below coded loads the R packages. 
```{r echo=TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

####The below code reads and processes the data, as well as prints the first few lines.  
Note, missing values are kept in the data.  
  
```{r echo=TRUE}
data <- read.csv("activity.csv", na.strings = "NA")
data$date <- as.Date(data$date,"%Y-%m-%d")
data_df <- tbl_df(data) %>%
        select(date, interval, steps)
data_df
```

## Part 2: What is mean total number of steps taken per day?
This section completes the following:

- Groups the data by day
- Plots a histogram of daily total steps
- Calculates and reports the mean and median number of steps per day  


####The  below code groups and summarizes the total steps by day, and plots a histogram of total daily steps.
```{r echo=TRUE}
day_totals <- data_df %>%
        group_by(date) %>%
        summarize(total_steps = sum(steps, na.rm = TRUE))

g <- ggplot(day_totals, aes(total_steps))
g + geom_histogram(bins = 50, fill = "dark red") + labs(x = "Daily Step Total")
```


####The below code prints the mean and median total steps per day. 
```{r echo=TRUE}
mean_daytotal <- mean(day_totals$total_steps)
median_daytotal <- median(day_totals$total_steps)
paste("The mean total steps per day is ", round(mean_daytotal,0),", and the median total steps per day is ", round(median_daytotal,0),".", sep = "")
```

## Part 3: What is the average daily pattern?
This section completes the following (see code below):

- Produces a time series plot of the average steps by time interval
- Reports the time interval that has the highest step count on average

```{r echo=TRUE}
interval_avg <- data_df %>%
        group_by(interval) %>%
        summarize(avg_steps = mean(steps, na.rm = TRUE))

g <- ggplot(interval_avg, aes(interval,avg_steps))
g + geom_line(color = "navy") + labs(y = "Average number of steps")

```

## Part 4: Imputing Missing Values
The section completes the following:

- Reports the number of missing values in the dataset
- Fills in missing values with an expected amount based on interval averages.
- Plots new histogram based on the revised data with missing values filled in  


####The below code determines the number of missing values in the data.
```{r echo=TRUE}
paste("The number of missing values is ", sum(is.na(data_df[,3])),".", sep = "")
```


####The below code imputes missing values (based on interval averages), and plots a new hostogram of daily step totals.
```{r echo=TRUE}
interval_means <- data_df %>%
        group_by(interval) %>%
        summarize(Avg_steps = mean(steps, na.rm = TRUE))

data_fill <- merge(data_df,interval_means, by = "interval") %>%
        arrange(date, interval) %>%
        mutate(steps_fill = ifelse(is.na(steps),Avg_steps,steps)) %>%
        select(date, interval, steps_fill) %>%
        tbl_df()

day_totals <- data_fill %>%
        group_by(date) %>%
        summarize(total_steps = sum(steps_fill, na.rm = TRUE))

library(ggplot2)
g <- ggplot(day_totals, aes(total_steps))
g + geom_histogram(bins = 50, fill = "dark red") + labs(x = "Daily Step Total")
```

####The below code prints the mean and median total steps per day, for the "imputed" dataset. 
```{r echo=TRUE}
mean_daytotal <- mean(day_totals$total_steps)
median_daytotal <- median(day_totals$total_steps)
paste("The mean total steps per day is now ", round(mean_daytotal,0),", and the median total steps per day is now ", round(median_daytotal,0),".", sep = "")
```

## Part 5: Activity patterns between weekdays and weekends
This section completes the following (see code below):

- Adds a weekday/weekend indicator to the modified "imputed" dataset
- Produces a time series panel plot of the average steps by time interval, for weekdays vs weekends
- Reports the time interval that has the highest step count on average

```{r echo=TRUE}
interval_avg <- data_fill %>%
        mutate(weekend = ifelse((weekdays(date) == "Saturday")|(weekdays(date) == "Sunday"),"weekend","weekday")) %>%
        group_by(weekend, interval) %>%
        summarize(avg_steps = mean(steps_fill))

g <- ggplot(interval_avg, aes(interval,avg_steps))
g + geom_line(color = "navy") + facet_wrap(~weekend,ncol = 1) + labs(y = "Average number of steps")

```


####The below code determines the intervals with highest average step count for weekdays vs weekends. 
```{r echo=TRUE}
highest <- data_fill %>%
        mutate(weekend = ifelse((weekdays(date) == "Saturday")|(weekdays(date) == "Sunday"),"weekend","weekday")) %>%
        group_by(weekend, interval) %>%
        summarize(avg_steps = mean(steps_fill)) %>%
        group_by(weekend) %>%
        summarize(highest = max(avg_steps))

interval_avg[interval_avg$avg_steps %in% highest$highest,]
```